---
title: "Seamless Integration of Mass Spectrometry Data from Different Sources with the `Spectra` Package"
author: "Johannes Rainer^[Institute for Biomedicine, Eurac Research, Bolzano,
Italy; johannes.rainer@eurac.edu], Sebastian Gibb^[], Laurent Gatto^[]"
output:
  rmarkdown::html_document:
    highlight: pygments
    toc: true
    toc_depth: 3
    fig_width: 5
vignette: >
  %\VignetteIndexEntry{Seamless Integration of Mass Spectrometry Data from Different Sources}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding[utf8]{inputenc}
  %\VignettePackage{SpectraTutorials}
  %\VignetteDepends{Spectra,mzR,MsBackendHmdb,MsBackendMgf,pheatmap,BiocStyle}
bibliography: references.bib
---

```{r style, echo = FALSE, results = 'asis', message = FALSE}
library(BiocStyle)
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

**Last modified:** `r file.info("analyzing-MS-data-from-different-sources-with-Spectra.Rmd")$mtime`<br />
**Compiled**: `r date()`


# Abstract

Mass spectrometry (MS) data is a key technology in modern proteomics and
metabolomics experiments. Due to continuous improvements in MS instrumentation,
the generated data can easily become very large. Also, different additional
resources of MS data exist, such as spectra libraries and databases, all with
their own specific file formats that sometimes do not support manipulations of
the original data.

Learning from experiences with the `r Biocpkg("MSnbase")` Bioconductor package
we developed a novel infrastructure to handle MS spectral data in R, the `r
Biocpkg("Spectra")` package. This package implements a clear separation of user
functionality from code to provide, store and import mass spectrometry
data. Different *backends* can hence be used that enable access to data from
various data resources or that are designed specifically for very large MS data
sets. Data manipulations are by default not directly applied to the data but
cached in a *lazy processing queue* which allows analyses also of *read-only*
data representations.

This workshop shows the expandability of the new infrastructure to enable a
seamless integration and analysis of MS data from a variety of input formats
illustrated by a simple matching of experimental MS2 spectra against a public
spectral database and export of the data in a format commonly used for exchange
of MS2 data.


# Installation

The source code of this tutorial is available at
[https://github.com/jorainer/SpectraTutorials](https://github.com/jorainer/SpectraTutorials). A
recent version or [R](https://r-project.org) (4.0 or higher) and a set of R and
Bioconductor packages are required to run the tutorial. These can be installed
with the R code shown below.

```{r installation, eval = FALSE}
install.packages(c("devtools", "rmarkdown", "BiocManager"))
BiocManager::install(c("BiocStyle",
                       "MsCoreUtils",
                       "Spectra",
                       "pheatmap"))
BiocManager::install("RforMassSpectrometry/MsBackendHmdb")
BiocManager::install("RforMassSpectrometry/MsBackendMgf")
```

To run the tutorial locally:
- clone [this github
repository](https://github.com/jorainer/SpectraTutorials), e.g. with `git clone
https://github.com/jorainer/SpectraTutorials`.
- to run also the code to import the MS2 spectra from HMDB the *All Spectra Files (XML)* from the [hmdb downloads
  page](https://hmdb.ca/downloads) has to be downloaded. The contents of the
  *hmdb_all_spectra.zip* archive should then be unzipped into the folder
  *data/hmdb_all_spectra*.

Alternatively, a [docker](https://www.docker.com/) image with all necessary
packages pre-installed is available
[here](https://hub.docker.com/r/jorainer/spectra_tutorials). This can be
installed with `docker pull jorainer/spectra_tutorials:latest`.

To run the docker use:

```r
docker run \
 	-e PASSWORD=bioc \
 	-p 8787:8787 \
 	jorainer/spectra_tutorials:latest
```

Interaction with the R within the running docker container is then possible
by entering the address `http://localhost:8787/` in a web browser and logging in
with user `rstudio` password `bioc`. This gives access to a RStudio instance
running within the container.


# Analysing MS data with `Spectra`

The `Spectra` package implements a clear separation of user functionality from
code to provide, store and read mass spectrometry data. Thus, different data or
file format-specific *backends* can be implemented and directly *plugged-in*
without affecting the way the user would access or analyze the data. This
represents an extension to the *in-memory* and *on-disk* data modes in the `r
Biocpkg("MSnbase")` package that enabled either fast data processings or
analysis of very large data sets by keeping only a limited amount of data in the
computer's memory [@gattoMSnbaseEfficientElegant2020a].

In this workshop we will import MS data from mzML files, match the MS2 fragment
spectra for one ion against MS2 spectra from a public database (i.e. the Human
Metabolome Database [HMDB](https://hmdb.ca)) and export the data as a MGF
file. For each of the data sources and the export file format we will use a
specific backend.

Below we import the MS data from the mzML files provided within this
package. These files contain MSn data of a mix of 8 standard compounds (solved
either in water or a pool of human serum samples) measured with a HILIC-based
LC-MS/MS setup. MS2 data was generated by data dependent acquisition using two
different collision energies. For data import and representation we use the
`MsBackendMzR` backend which supports import (and export) of data from the most
common *raw* mass spectrometry file formats (i.e. mzML, mzXML and CDF).

```{r data-import}
library(Spectra)
fls <- dir(system.file("mzML", package = "SpectraTutorials"),
           full.names = TRUE)
sps_all <- Spectra(fls, backend = MsBackendMzR())
```

The MS data is represented by a `Spectra` object, which can be thought of as
being a `data.frame` containing MS data, columns being the spectra variables
(such as `"rtime"`, i.e. the retention time) and rows the individual spectra.
Individual spectra variables can be accessed either via `$` and their name or
using their dedicated access functions (which is the preferred way). Below we
access the retention times of the first spectra using either `rtime` or
`$rtime`. Note that we could use the `spectraVariables` function on our object
to get the full listing of all available spectra variables.

```{r rtime}
head(rtime(sps_all))
head(sps_all$rtime)
```

Our `Spectra` object contains information from in total `r length(sps_all)`
spectra from `length(unique(dataOrigin(sps_all)))` mzML files. By using the
`MsBackendMzR` backend only general information about each spectrum is kept in
memory resulting in a low memory footprint.

```{r on-disk-size}
print(object.size(sps_all), units = "MB")
```

We can also load the full data into memory by changing the backend from
`MsBackendMzR` to `MsBackendDataFrame`. This does not affect the way how we
can use the `Spectra` object itself. The same operations and functions are
available to us, independently of the way how the data is stored (i.e. which
backend is used).

```{r change-backend}
sps_all <- setBackend(sps_all, MsBackendDataFrame())
```

The size of our `Spectra` object is now however larger, since the full data has
been loaded into memory.

```{r in-mem-size}
print(object.size(sps_all), units = "MB")
```

Next we subset our data to MS2 spectra with a precursor ion that matches the m/z
of the [M+H]+ ion of the metabolite Cystine (accepting a difference of 10
parts-per-million (ppm)).

```{r}
mz <- 241.0311

sps <- filterPrecursorMz(sps_all, mz = mz + ppm(c(-mz, mz), 10))
sps
```

We could identify `r length(sps)` spectra matching our condition. The plot below
shows the first spectrum.

```{r raw-ms2, fig.width = 7, fig.height = 7}
plotSpectra(sps[1])
```

The spectrum contains still very low abundance peaks, most likely representing
noise. Thus we next filter the spectra removing all peaks with an intensity
smaller 5% of the maximum intensity of each spectrum. To this end we define a
function that takes intensity values from each spectrum and returns a logical
value whether the peak should be kept (`TRUE`) or not (`FALSE`). This function
is passed along to the `filterIntensity` function.

```{r filter-intensity}
low_int <- function(x, ...) {
    x > max(x, na.rm = TRUE) * 0.05
}
sps <- filterIntensity(sps, intensity = low_int)
```

In addition we *normalize* each spectrum replacing the absolute intensity values
to relative values related to the maximum intensity (which is set to 100). For
this operation we also define a function which is applied to the peak matrix of
each spectrum by passing it as parameter `FUN` to the `addProcessing` function.

```{r normalize}
norm_int <- function(x, ...) {
    maxint <- max(x[, "intensity"], na.rm = TRUE)
    x[, "intensity"] <- 100 * x[, "intensity"] / maxint
    x
}
sps <- addProcessing(sps, norm_int)
```

To show the effect of the normalization we extract the intensities of the first
spectrum:

```{r}
intensity(sps)[[1]]
```

We next perform a pairwise comparison of the spectra using the dot product as
the similarity measure to evaluate whether all spectra are the same or
represent different fragmentations. Prior to the similarity calculation the
peaks of the individual spectra have to be matched against each other (based on
their m/z). We specify `ppm = 20` so that peaks with a difference in m/z smaller
than 20ppm will be considered matching.

```{r}
cormat <- compareSpectra(sps, ppm = 20)
```

The similarity of the spectra with each other is represented in the heatmap
below.

```{r ms2-heatmap, fig.width = 7, fig.height = 7}
library(pheatmap)
hm <- pheatmap(cormat, cutree_rows = 3)
```

The spectra group into 3 clusters, which are related to the collision energy
used for the fragmentation (see below; the collision energy is encoded in the
file name as CE20 and CE30). We subsequently reduce our dataset to the cluster
with the spectra generated with a collision energy of 20.

```{r}
split(basename(dataOrigin(sps)), cutree(hm$tree_row, 3))
sps_ce20 <- split(sps, cutree(hm$tree_row, 3))[[1L]]
```

Although the precursor m/z of our spectra matches the m/z of Cystine, we can not
be sure that they might not represent fragmentation of ions from other
compounds. We thus aim to compare our experimental MS2 spectra against spectra
from a public database, for our example the Human Metabolome Database
(HMDB). Predicted and experimental MS2 spectra are provided by HMDB as
individual xml files in a custom file format which are bundled (and can hence be
downloaded) in a single archive. The `MsBackendHmdbXml` backend (provided by the
[MsBackendHmdb](https://github.com/RforMassSpectrometry/MsBackendHmdb) package)
allows to import spectral data from these xml files. To reproduce the following
code it is expected (as detailed in the Installation section) that all xml files
from HMDB are available in a folder *data/hmdb_all_spectra*. Below we identify
all xml files containing the key word `"ms_ms"` in their file name and load them
into a `Spectra` object using the `MsBackendHmdbXml` backend. Note that this
import operation from ~ 500,000 individual xml files takes a long time.

```{r hmdb-import, eval = FALSE}
library(MsBackendHmdb)
fls <- dir("data/hmdb_all_spectra/", full.names = TRUE, pattern = "ms_ms")
hmdb <- Spectra(fls, source = MsBackendHmdbXml(), nonStop = TRUE)
```

```{r hmdb-import-save, eval = FALSE, echo = FALSE}
## Eventually save for faster re-use
save(hmdb, file = "data/hmdb.RData", compress = "xz")
```

```{r hmdb-load, echo = FALSE}
library(MsBackendHmdb)
data("hmdb", package = "SpectraTutorials")
```

With this we have now also `Spectra` object containing all MS2 spectra from
HMDB.

```{r}
hmdb
```

To avoid comparing our experimental spectra against all these ~500,000 spectra,
we first determine using the `containsMz` function which of the HMDB spectra
contain a peak matching the m/z of our ion of interest. We have to use a rather
large `tolerance` value (which defines the maximal acceptable absolute
difference in m/z values) since some of the experimental spectra in HMDB seem to
be recorded by not well calibrated instruments.

```{r has-mz}
has_mz <- containsMz(hmdb, mz = mz, tolerance = 0.2)
```

In total `r sum(has_mz)` spectra contain a peak with the required m/z (+/- 0.2
dalton) and we can proceed to calculate spectral similarities between our
experimental spectra and this subset from HMDB.

```{r compare-subset}
hmdb_with_mz <- hmdb[has_mz]
res <- compareSpectra(sps_ce20, hmdb_with_mz, tolerance = 0.2)
```

The highest similarity between our spectra and the spectra from HMDB is `r
max(res)`. Below we compare the two best matching spectra with a *mirror plot*,
in the upper panel showing our experimental spectrum and in the lower panel the
best matching MS2 spectrum from HMDB.

```{r mirror-plot, fig.width = 7, fig.height = 7}
idx <- which(res == max(res), arr.ind = TRUE)
## Specifying a function to draw peak labels
label_fun <- function(x) {
    format(unlist(mz(x)), digits = 4)
}
plotSpectraMirror(sps_ce20[idx[1]], hmdb_with_mz[idx[2]], tolerance = 0.2,
                  labels = label_fun, labelPos = 2, labelOffset = 0.2,
                  labelSrt = -30)
grid()
```

Our experimental spectrum seems to nicely match the *reference* MS2 in
HMDB. Below we extract its compound identifier from the HMDB spectrum (stored in
a spectra variable called `"compound_id"`) and add it as an additional spectra
variable `"hmdb_id"` to our experimental spectra. We also set the collision
energy for these spectra to 20eV.

```{r}
sps_ce20$hmdb_id <- hmdb_with_mz[idx[2]]$compound_id
sps_ce20$collisionEnergy <- 20
```

In fact, the matching spectrum from HMDB is an experimental spectrum for
[L-Cystine](https://hmdb.ca/metabolites/HMDB0000192).

At last we want to export our spectra to a file in MGF format. The `Spectra`
package does however not have any built-in support for this file format. The
required functionality is implemented in the
[MsBackendMgf](https://github.com/RforMassSpectrometry/MsBackendMgf) R package
and can thus be added to `Spectra` by using `MsBackendMgf` backend provided by
that package.

Data from `Spectra` objects can be exported with the `export` function. The
format in which the data is exported depends on the specified `MsBackend`
class. By using an instance of `MsBackendMgf` we can write the data to a file in
MGF format.

```{r export}
library(MsBackendMgf)
export(sps_ce20, backend = MsBackendMgf(), file = "Cystine_ce20.mgf")
```


# Summary

With the simple use case of matching experimental MS2 spectra against a public
database we illustrated in this short tutorial the flexibility and expandability
of the `Spectra` package that enables the seamless integration of mass
spectrometry data from different sources. This was only possible with a clear
separation of the user functionality (`Spectra` object) from the representation
of the data (`MsBackend` object). Backends such as the
[`MsBackendMgf`](https://github.com/RforMassSpectrometry/MsBackendMgf) or
[`MsBackendHmdbXml`](https://github.com/RforMassSpectrometry/MsBackendHmdb) can
provide support for additional data formats or data sources, while others, due
to their much lower memory footprint (`MsBackendMzR`, `MsBackendHdf5Peaks`),
enable the analysis of also very large data sets. Most importantly however, these
backends are interchangeable and do not affect the way users can handle and
analyze MS data with the `Spectra` package.


# References
