---
title: "MS/MS Spectra Matching with the `MetaboAnnotation` Package"
author: "Johannes Rainer^[Institute for Biomedicine, Eurac Research, Bolzano, Italy; johannes.rainer@eurac.edu], Michael Witting^[Research Unit Analytical BioGeoChemistry, Helmholtz Zentrum München and Chair of Analytical Food Chemistry, TUM School or Life Sciences, Technical University of Munich, Germany]"
output:
  rmarkdown::html_document:
    highlight: pygments
    toc: true
    toc_depth: 3
    fig_width: 5
vignette: >
  %\VignetteIndexEntry{MS/MS Spectra Matching with the `MetaboAnnotation` Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding[utf8]{inputenc}
  %\VignettePackage{SpectraTutorials}
  %\VignetteDepends{Spectra,mzR,RMariaDB,BiocStyle,MetaboAnnotation,pander,msdata,MsBackendMassbank}
bibliography: references.bib
---

```{r style, echo = FALSE, results = 'asis', message = FALSE}
library(BiocStyle)
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

**Last modified:** `r file.info("Spectra-matching-with-MetaboAnnotation.Rmd")$mtime`<br />
**Compiled**: `r date()`

# Overview

## Introduction


The `Spectra` package provides all the functionality required for annotation and
identification workflows for untargeted LC-MS/MS data, but, while being very
flexible and customizable, it might be too cumbersome for beginners or analysts
not accustomed with R. To fill this gap we developed the
[`MetaboAnnotation`](https://rformassspectrometry.github.io/MetaboAnnotation)
package that builds upon `Spectra` and provides functions for annotation of
LC-MS and LC-MS/MS data sets tailored towards the less experienced R user
[@rainer_modular_2022].

## Convenient spectra matching using `MetaboAnnotation`

In this example use case we match experimental MS2 spectra from a DDA experiment
on a pesticide mix against reference spectra from MassBank. Below we load the
experimental data file which is distributed *via* the `msdata` R package.

```{r}
library(Spectra)
library(pander)

#' Load the pesticide mix data
fl <- system.file("TripleTOF-SWATH", "PestMix1_DDA.mzML", package = "msdata")
pest <- Spectra(fl)
```

We next restrict the data set to MS2 spectra only and in addition *clean* these
spectra by removing all peaks from a spectrum that have an intensity lower than
5% of the largest peak intensity of that spectrum. Finally, single-peak spectra
are removed.

```{r}
#' restrict to MS2 data and remove intensities with intensity lower 5%
pest <- filterMsLevel(pest, msLevel = 2L)

#' Remove peaks with an intensity below 5% or the spectra's BPC
low_int <- function(x, ...) {
    x > max(x, na.rm = TRUE) * 0.05
}
pest <- filterIntensity(pest, intensity = low_int)

#' Remove peaks with a single peak
pest <- pest[lengths(pest) > 1]

```

This leads to a data set consisting of `r length(pest)` spectra. We next connect
to a MassBank database (running within this docker image) and create a `Spectra`
object representing that data.

```{r massbank-library}
library(RMariaDB)
library(MsBackendMassbank)

#' Connect to the MassBank MySQL database
con <- dbConnect(MariaDB(), user = "massbank", dbname = "MassBank",
                 host = "localhost", pass = "massbank")
mbank <- Spectra(con, source = MsBackendMassbankSql())
```

Alternatively, if no MySQL database system is available or if this tutorial can
not be run within *docker*, an SQLite database version of the MassBank data (in
the format required for `MsBackendMassbankSql` is available
[here](https://github.com/jorainer/SpectraTutorials/releases/tag/2021.03). After
downloading the database *MassbankSql-2021-03.db* to the current R workspace,
the resource can be loaded with:

```{r massbank-sqlite, eval = FALSE}
## Alternative to the MySQL server - assuming the SQLite database was
## stored to the R working directory.
library(RSQLite)
con <- dbConnect(SQLite(), "MassbankSql-2021-03.db")
mbank <- Spectra(con, source = MsBackendMassbankSql())
```

We could now directly calculate similarities between the `r length(pest)`
experimental (query) MS2 spectra and the `r length(mbank)` MassBank reference
(target) spectra using the `compareSpectra` method, but this would be
computationally very intense because a similarity score would be calculated
between each query and each target spectrum. As alternative we use here the
`matchSpectra` function from the
[`MetaboAnnotation`](https://github.com/RforMassSpectrometry/MetaboAnnotation)
package that allows to restrict similarity calculations between query and target
spectra with similar m/z of their precursor ion or have a similar retention
time.

Below we create a `CompareSpectraParam` object setting parameter
`requirePrecursor = TRUE` (to restrict similarity calculations only to query and
target spectra with a similar precursor m/z) and `ppm = 10` (m/z difference
between the query and target precursor has to be within 10 ppm). Parameter
`THRESHFUN` enables to define a *threshold function* that defines which spectra
are considered matching. With the function used below only MS2 spectra with a
similarity (calculated with the default *dotproduct* function) larger or equal
to 0.8 are considered matching.

```{r}
library(MetaboAnnotation)
prm <- CompareSpectraParam(ppm = 10, requirePrecursor = TRUE,
                           THRESHFUN = function(x) which(x >= 0.8))
```

We next call `matchSpectra` with this parameter object and pass `pest` and
`mbank` as query and target `Spectra`, respectively. This takes approximately 1
minute to complete, which is not tremendously fast, but still much faster than a
pairwise comparison between all query and target spectra would be.

```{r}
mtch <- matchSpectra(pest, mbank, param = prm)
mtch
```

As a result we get a `MatchedSpectra` object that contains the query and target
spectra as well as the matching result (i.e. the information which query
spectrum matches with which target spectrum based on what similarity score). We
can use the `query` and `target` functions to access the query and target
`Spectra` objects and `matches` to extract the matching information. Below we
display the first 6 rows of that matrix.

```{r}
head(matches(mtch))
```

Functions `whichQuery` and `whichTarget` return the (unique) indices of the
query and target spectra that could be matched.

```{r}
whichQuery(mtch)
```

As we can see only few of the query spectra (`r length(whichQuery(mtch))` of the
`r length(query(mtch))` spectra) could be matched. This is in part because for a
large proportion spectra in MassBank no precursor m/z is available and with
`requirePrecursor = TRUE` these are not considered in the similarity
calculation. Setting `requirePrecursor = FALSE` would calculate a similarity
between all spectra (even those with missing precursor information) but
calculations can take up to several hours.

```{r}
sum(is.na(precursorMz(mbank)))
```

The `MatchedSpectra` object inherits much of the functionality of a `Spectra`
object. `spectraVariables` returns for example all the available *spectra
variables*, from both the query as well as the target `Spectra`. The variable
names of the latter are prefixed with `target_` to discriminate them from the
variable names of the query.

```{r}
spectraVariables(mtch)
```

We can access spectra individual variables using `$` and the variable name, or
multiple variables with the `spectraData` function. Below we extract the
retention time, the precursor m/z of the query spectrum, the precursor m/z of
the target spectrum as well as the similarity score from the object using the
`spectraData` function.

```{r}
spectraData(mtch, c("rtime", "precursorMz", "target_precursorMz", "score"))
```

The returned `DataFrame` contains the matching information for the full data
set, i.e. of each query spectrum and hence, returns `NA` values for query
spectra that could not be matched with a target spectrum. Note also that query
spectra matching multiple target spectra are represented by multiple rows (one
for each matching target spectrum).

Here we're only interested in query spectra for which a match was found and thus
we subset the `MatchedSpectra` to query spectra with a matching target spectrum.

```{r}
mtch <- mtch[whichQuery(mtch)]
```

Subsetting of `MatchedSpectra` is always relative to the query, i.e. subsetting
an object with an index `4` would restrict the object to only the matching
results for the 4th query spectrum.

We now extract the matching information after subsetting:

```{r}
spectraData(mtch, c("rtime", "precursorMz", "target_precursorMz", "score"))
```

We can also return the compound names for the matching spectra.

```{r, results = "asis"}
pandoc.table(style = "rmarkdown",
    as.data.frame(spectraData(mtch, c("rtime", "target_compound_name",
                                      "score"))))
```

The `matchSpectra` enables thus to perform convenient spectra matching between
MS data represented as `Spectra` objects. As a result, a `MatchedSpectra` object
is returned that, in addition to the matching results, contains also the query
and target spectra. Pre-filtering the spectra prior to the actual spectra
similarity calculation can reduce the running time of a `matchSpectra` call but
might also miss some potential matches. Note that in addition to the precursor
m/z-based pre-filter also retention time pre-filtering would be available (see
`?matchSpectra` for more information). Also, a more advanced matching approach
would be available with the `MatchForwardReverseParam` that calculates in
addition to the *forward score* also a *reverse similarity* for each match.


## MS2 spectra matching in an `xcms` workflow

In LC-MS/MS-based untargeted metabolomics (or small compound mass spectrometry
experiments in general) quantification of the compounds is performed in MS1
while the MS2 data is used for identification of the features. Quantification of
MS1 data requires a chromatographic peak detection step which can be performed
using the functionality from the `r BiocStyle::Biocpkg("xcms")` package. Below
we load thus the `xcms` package and import the full MS data using the
`readMSData` function.

```{r}
library(xcms)
pest_all <- readMSData(fl, mode = "onDisk")
```

We next perform the chromatographic peak detection using the *centWave*
algorithm (see the *LC-MS/MS data analysis with xcms* vignette from the `xcms`
package for details on the chromatographic peak detection settings).

```{r}
cwp <- CentWaveParam(snthresh = 5, noise = 100, ppm = 10,
                     peakwidth = c(3, 30))
pest_all <- findChromPeaks(pest_all, param = cwp)
```

In total `r nrow(chromPeaks(pest_all))` chromatographic peaks have been
identified. Below we display the first 6 of them.

```{r}
head(chromPeaks(pest_all))
```

We can now extract all MS2 spectra for each chromatographic peak with the
`chromPeakSpectra` function. This function identifies all MS2 spectra recorded
by the instrument with a retention time within the retention time and with a
precursor m/z within the m/z boundaries of the chromatographic peak. By setting
`return.type = "Spectra"` we ensure that the data is being returned in the newer
`Spectra` format hence enabling the simplified spectra matching with the
functionality presented here.

```{r}
pest_ms2 <- chromPeakSpectra(pest_all, return.type = "Spectra")
pest_ms2
```

Spectra variable `peak_id` contains the identified of the chromatographic peak
(i.e. its row name in `chromPeaks`).

```{r}
pest_ms2$peak_id
```

We next, like in the previous section, *clean up* the spectra removing peaks
with an intensity below 5% of the largest peak intensity per spectrum and
removing spectra with a single peak.

```{r}
#' Remove peaks with an intensity below 5%
pest_ms2 <- filterIntensity(pest_ms2, intensity = low_int)

#' Remove peaks with a single peak
pest_ms2 <- pest_ms2[lengths(pest_ms2) > 1]
```

In addition we scale also the intensities within each MS2 spectrum by replacing
them with intensities relative to the maximum peak intensity (see [here](https://jorainer.github.io/SpectraTutorials/articles/analyzing-MS-data-from-different-sources-with-Spectra.html#data-processing-and-manipulation-1) for
more information). In addition to the query spectra, we also *normalize* the
MassBank spectra in the same way.

```{r normalize}
#' Define a function to *normalize* the intensities
norm_int <- function(x, ...) {
    maxint <- max(x[, "intensity"], na.rm = TRUE)
    x[, "intensity"] <- 100 * x[, "intensity"] / maxint
    x
}
#' *Apply* the function to the data
pest_ms2 <- addProcessing(pest_ms2, norm_int)
mbank <- addProcessing(mbank, norm_int)
```

Next we perform now the spectra matching with the same parameters as in the
previous section.

```{r}
pest_match <- matchSpectra(pest_ms2, mbank, param = prm)
pest_match
```

Again, we restrict the `MatchedSpectra` to query spectra which could be matched.

```{r}
pest_match <- pest_match[whichQuery(pest_match)]
```

The table below lists the compound names of matching spectra for the
chromatographic peaks.

```{r, results = "asis"}
pandoc.table(
    style = "rmarkdown",
    as.data.frame(spectraData(pest_match, c("peak_id", "rtime",
                                            "target_compound_name"))))
```

We can also directly plot matching (query and target) spectra against each other
using the `plotSpectraMirror` function subsetting the `MatchedSpectra` object
to the query spectrum of interest. Below we plot the third query spectrum
against all of its matching target spectra.

```{r}
plotSpectraMirror(pest_match[3])
```

Summarizing, with the `chromPeakSpectra` and the `featureSpectra` functions,
`xcms` allows to return MS data as `Spectra` objects which enables, as shown in
this simple example, to perform MS2 spectra matching using the `Spectra` as well
as the `MetaboAnnotation` packages hence simplifying MS/MS-based annotation of
LC-MS features from `xcms`.


# Session information

```{r}
sessionInfo()
```

# References
