---
title: "`Spectra`: an expandable infrastructure to handle mass spectrometry data"
author: "Johannes Rainer^[Institute for Biomedicine, Eurac Research, Bolzano, Italy; johannes.rainer@eurac.edu], Michael Witting^[Research Unit Analytical BioGeoChemistry, Helmholtz Zentrum München and Chair of Analytical Food Chemistry, TUM School or Life Sciences, Technical University of Munich, Germany], Sebastian Gibb^[Department of Anaesthesiology and Intensive Care, University Medicine Greifswald, Germany], Laurent Gatto^[Computational Biology Unit, de Duve Institute, UCLouvain, Brussels, Belgium]"
output:
  rmarkdown::html_document:
    highlight: pygments
    toc: true
    toc_depth: 3
    fig_width: 5
vignette: >
  %\VignetteIndexEntry{`Spectra`: an expandable infrastructure to handle mass spectrometry data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding[utf8]{inputenc}
  %\VignettePackage{SpectraTutorials}
  %\VignetteDepends{Spectra,mzR,BiocStyle,msdata,MsBackendSql,RSQLite,microbenchmark}
bibliography: references.bib
---

```{r style, echo = FALSE, results = 'asis', message = FALSE}
library(BiocStyle)
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

**Last modified:** `r file.info("Spectra-backends.Rmd")$mtime`<br />
**Compiled**: `r date()`


# Abstract

Mass spectrometry (MS) data is a key technology in modern metabolomics and
proteomics experiments. Continuous improvements in MS instrumentation, larger
experiments and new technological developments lead to ever growing data sizes
and increased number of available variables making standard in-memory data
handling and processing difficult.

The `r BiocStyle::Biocpkg("Spectra")` package provides a modern infrastructure
for MS data handling specifically designed to enable extension to additional
data resources or alternative data representations. These can be realized by
extending the virtual `MsBackend` class and its related methods. Implementations
of such `MsBackend` classes can be tailored for specific needs, such as low
memory footprint, fast processing, remote data access, or also support for
specific additional data types or variables. Importantly, data processing of
`Spectra` objects is independent of the backend in use due to a lazy evaluation
mechanism that caches data manipulations internally.

This workshop discusses different available data representations for MS data
along with their properties, advantages and performances. In addition,
*Spectra*'s concept of lazy evaluation for data manipulations is presented, as
well as a simple caching mechanism for data modifications. Finally, it explains
how new `MsBackend` instances can be implemented and tested to ensure
compliance.


# Introduction

This workshop/tutorial assumes that readers are familiar with mass spectrometry
data. See the [*LC-MS/MS in a
nutshell*](https://jorainer.github.io/SpectraTutorials/articles/analyzing-MS-data-from-different-sources-with-Spectra.html#lc-msms-in-a-nutshell)
section of the *Seamless Integration of Mass Spectrometry Data from Different
Sources* vignette in this package for a general introduction to MS.


## Pre-requisites

- Basic familiarity with R and Bioconductor.
- Basic understanding of Mass Spectrometry (MS) data.


## Installation and Participation

- Get the [docker image](https://hub.docker.com/r/jorainer/spectra_tutorials) of
  this tutorial with `docker pull jorainer/spectra_tutorials:latest`.
- Start docker using
  ```
  docker run \
      -e PASSWORD=bioc \
      -p 8787:8787 \
      jorainer/spectra_tutorials:latest
  ```
- Enter `http://localhost:8787` in a web browser and log in with username
  `rstudio` and password `bioc`.
- Open this R-markdown file
  (*vignettes/Spectra-backends.Rmd*) in the
  RStudio server version in the web browser and evaluate the R code blocks.
- To get the source code: clone [this github
repository](https://github.com/jorainer/SpectraTutorials), e.g. with `git clone
https://github.com/jorainer/SpectraTutorials`.

## R/Bioconductor packages used

- [`Spectra`](https://bioconductor.org/packages/Spectra)
- [`MsBackendSql`](https://bioconductor.org/packages/MsBackendSql)


## Workshop goals and objectives

This is a more technical demonstration of the internals of the *Spectra* package
and design of its MS infrastructure. We're not demonstrating any use cases or
analysis workflows here.


### Learning goals

- Understand how MS data is handled with *Spectra*.
- Understand differences and properties of different `MsBackend`
  implementations.


### Learning objectives

- Learn how MS data is handled with `Spectra`.
- Understand which data representations/backends fit which purpose/use case.
- Insights into the internals of the `Spectra` MS infrastructure to facilitate
  implementation of own backend(s).


# Workshop

## The `Spectra` package

- **Purpose**: provide an expandable, well tested and user-friendly
  infrastructure for mass spectrometry (MS) data.
- **Design**: separation between user interface (functions) and code to provide,
  store and read MS data.

![Spectra: separation into user functionality and data
representation](Spectra.png)

- `Spectra`: provides the interface for the user.
- `MsBackend`: defines how and where data is stored and how it is managed.
- **Why?**:
  - user does not need to care about where or how the data is stored.
  - the same functionality can be applied to MS data, regardless of the used
    backend.
  - enables specialized data storage/representation options: remote data, low
    memory footprint, high performance.


## Creating and using `Spectra` objects

MS data is consists in general of duplets of mass-to-charge (*m/z*) and
intensity values along with potential additional information, such as the MS
level, retention time, polarity etc. In its easiest form, MS data consists thus
of two (aligned) numeric vectors with the m/z and intensity values and an
e.g. `data.frame` containing the additional annotations for the spectrum. In
`Spectra` terminology, `"mz"` and `"intensity"` are called *peak variables*
(because they provide information on individual mass peaks), and all other
additional annotations *spectra variables*. Below we define m/z and intensity
values for a mass spectrum as well as a `data.frame` with additional *spectra
variables*.

```{r}
#' Define simple MS data
mz <- c(56.0494, 69.0447, 83.0603, 109.0395, 110.0712,
        111.0551, 123.0429, 138.0662, 195.0876)
int <- c(0.459, 2.585, 2.446, 0.508, 8.968, 0.524, 0.974, 100.0, 40.994)

sv <- data.frame(msLevel = 1L, polarity = 1L)
```

This basically represents now MS data in R. For obvious reasons it makes more
sense to store such data in a single *container*. We thus below create a
`Spectra` object from this MS data. We need to pass a `data.frame` with the
full data (m/z, intensity and spectra variables) to the `Spectra` constructor,
each row in that `data.frame` representing one spectrum. Thus we need to add the
m/z and intensity values as a `list` to the data frame.

```{r}
#' Create a Spectra object
library(Spectra)
sv$mz <- list(mz)
sv$intensity <- list(int)
s <- Spectra(sv)
s
```

We have now created a `Spectra` object. Individual spectra variables can be
accessed with either `$` and the name of the spectra variable, or with one of
the dedicated accessor functions.

```{r}
#' Access the MS level spectra variable
s$msLevel
msLevel(s)
```

Also, while we provided only the spectra variables `"msLevel"` and `"polarity"`,
more variables are available by default. These are called *core spectra
variables* and they can **always** be extracted from a `Spectra` object, even if
they are not defined (in which case a `NA` is returned). The list of all
available spectra variables in a `Spectra` can be extracted with
`spectraVariables`:

```{r}
#' List all available spectra variables
spectraVariables(s)
```

And values for these can always be extracted from a `Spectra` object.

```{r}
#' Extract the retention time spectra variable
s$rtime
```

So, we've got now a `Spectra` object representing a single MS spectrum - but, as
the name of the class implies, it is actually designed to represent many
spectra. Below we thus define again a `data.frame`, this time with 3 rows and
create a `Spectra` object from that. Also, we define some additional spectra
variables providing the name and ID of the compounds the MS2 (fragment) spectrum
represents.

```{r}
#' Define spectra variables for 3 MS spectra.
sv <- data.frame(
    msLevel = c(2L, 2L, 2L),
    polarity = c(1L, 1L, 1L),
    id = c("HMDB0000001", "HMDB0000001", "HMDB0001847"),
    name = c("1-Methylhistidine", "1-Methylhistidine", "Caffeine"))

#' Assign m/z and intensity values.
sv$mz <- list(
    c(109.2, 124.2, 124.5, 170.16, 170.52),
    c(83.1, 96.12, 97.14, 109.14, 124.08, 125.1, 170.16),
    c(56.0494, 69.0447, 83.0603, 109.0395, 110.0712,
      111.0551, 123.0429, 138.0662, 195.0876))
sv$intensity <- list(
    c(3.407, 47.494, 3.094, 100.0, 13.240),
    c(6.685, 4.381, 3.022, 16.708, 100.0, 4.565, 40.643),
    c(0.459, 2.585, 2.446, 0.508, 8.968, 0.524, 0.974, 100.0, 40.994))

#' Create a Spectra from this data.
s <- Spectra(sv)
s
```

Now we have a `Spectra` object representing 3 mass spectra and a set of spectra
and peak variables.

```{r}
#' List available spectra and peaks variables.
spectraVariables(s)
peaksVariables(s)
```

The spectra are organized linearly, one after the other. Thus, our `Spectra` has
a length of 3 and we can also subset the object using `[`.

```{r}
length(s)

#' Extract the 2nd spectrum
s[2]
```

Thus, a `Spectra` *acts* similar to a `data.frame` with rows being individual
spectra and columns being the spectra (or peaks) variables, that can be accessed
with the `$` operator.

```{r}
s$msLevel
```

Similarly (as with a `data.frame`) we can also add new spectra variables
(columns) using `$<-`.

```{r}
#' Add a new spectra variable.
s$new_variable <- c("a", "b", "c")
spectraVariables(s)
```

Finally, we can also visualize the peaks data of a spectrum in the `Spectra`.

```{r}
#' Plot the second spectrum.
plotSpectra(s[2])
```

Experimental MS data is generally (ideally) stored in files in mzML, mzXML or
CDF format. These are open file formats that can be read by most software. In
Bioconductor, the `r Biocpkg("mzR")` package allows to read/write data in these
formats. Below we create a `Spectra` object for such experimental data from
two mzML files that are provided by Bioconductor's `r Biocpkg("msdata")`
package.

```{r}
#' Import MS data from 2 mzML files.
fls <- dir(system.file("sciex", package = "msdata"), full.names = TRUE)
s2 <- Spectra(fls, source = MsBackendMzR())
s2
```

We have thus access to the raw experimental MS data through this `Spectra`
object. Like before, we can subset the object or access any of its spectra
variables. Note that for this `Spectra` object we have more, and different
spectra variables available (depending on what is provided by the raw data
files).

```{r}
#' Available spectra variables
spectraVariables(s2)

#' Access the MS levels for the first 6 spectra
s2$msLevel |> head()
```

And we can again visualize the data.

```{r}
#' Plot the 4th spectrum.
plotSpectra(s2[4])
```


## Use of different *backends* with `Spectra`, why and how?

While both `Spectra` objects *behave* the same way, i.e. all data can be
accessed in the same way, they actually use different *backends*:

```{r}
class(s@backend)
class(s2@backend)
```

Our first example `Spectra` uses thus a `MsBackendMemory` backend that, as the
name says, keeps all MS data in memory (within the backend object). In contrast,
the second `Spectra` which represents the experimental MS data from the two mzML
files uses the `MsBackendMzR` backend. This backend loads only general spectra
data (variables) into memory while importing the full peaks data from the
original data files only on demand (similar to the *on-disk* mode discussed in
[@gattoMSnbaseEfficientElegant2020a]). The advantage of this *offline* storage
mode is a lower memory footprint that enables also the analysis of large data
experiments on *standard* computers.

Why different backends?

**Reason 1**: support for additional file formats. The backend class defines the
functionality to import/export data from/to specific file formats. The `Spectra`
class stays *data storage agnostic*. Support for additional file formats can be
implemented independently of the *Spectra* package, contributed by different
developers. Examples:
- `MsBackendMgf` (defined in `r Biocpkg("MsBackendMgf")`) adds support for MS
  files in MGF file format.
- `MsBackendMsp` (defined in `r Biocpkg("MsBackendMsp")` adds support for MSP
  files.
- `MsBackendRawFileReader` (defined in `r Biocpkg("MsBackendRawFileReader")`)
  adds support for MS data files in Thermo Fisher Scientific's raw data file
  format.

To import data using such alternative backends it needs to be provided with the
`source` parameter in the `Spectra` constructor call:

```{r}
#' Import MS data from mzML files
s_mzr <- Spectra(fls, source = MsBackendMzR())
```

**Reason 2**: support different implementations to store or represent the
data. This includes both *where* the data is stored or *how* it is
stored. Specialized backends can be defined that are for example optimized for
performance or for low memory demand. Examples:
- `MsBackendMemory`: keeps all the MS data in memory and is optimized for a fast
  and efficient access to the peaks data matrices (i.e., the *m/z* and intensity
  values of the spectra).
- `MsBackendMzR`: keeps only spectra variables in memory and retrieves the peaks
  data on-the-fly from the original data files upon request. This guarantees a
  lower memory footprint and enables also analysis of large MS experiments.
- `MsBackendSql` (defined in the `r Biocpkg("MsBackendSql")` package): all MS
  data is stored in a SQL database. Has minimal memory requirements but any data
  (whether spectra or peaks variables) need to be retrieved from the
  database. Depending on the SQL database system used, this backend would also
  allow remote data access.

To evaluate and illustrate the properties of these different backends we create
below `Spectra` objects for the same data, but using different
backends. We use for that the `setBackend` function that allows to change the
backend of a `Spectra` object.

```{r}
#' Change the backend to MsBackendMemory
s_mem <- setBackend(s_mzr, MsBackendMemory())
```

With this we loaded all MS data from the original data files into memory. As a
third option we next store the full MS data into a SQL database by
using/changing to a `MsBackendOfflineSql` backend defined by `r
Biocpkg("MsBackendSql")` package. For the `setBackend` call we need to provide
the connection information for the database that should contain the data. This
includes the database driver that should be used (parameter `drv`, depending on
the database system that should be used), the database name (parameter `dbname`)
as well as eventual additional connection information like the host, username,
port or password. Which of these parameters are required depend on the SQL
database used and hence the driver (see also `?dbConnect` for more information
on the parameters). In our example we will store the data in a simple small
SQLite database, thus we use `drv = SQLite()` and provide with parameter
`dbname` the name of the SQLite database file (that should not yet exist). In
addition, importantly, we disable parallel processing with `BPPARAM =
SerialParam()` since most SQL databases don't support parallel data import.

```{r, echo = FALSE, results = "hide"}
if (file.exists("ms_backend_sql_test.sqlite"))
    file.remove("ms_backend_sql_test.sqlite")
```

```{r}
#' Change the backend to a SQL representation of the data.
library(MsBackendSql)
library(RSQLite)
s_db <- setBackend(s_mzr, MsBackendOfflineSql(), drv = SQLite(),
                   dbname = "ms_backend_sql_test.sqlite",
                   BPPARAM = SerialParam())
```

*Note*: a more efficient way to import MS data from data files into a SQL
database is the `createMsBackendSqlDatabase` from the *MsBackendSql*
package. Also, for larger data sets it is suggested to use more advanced SQL
database systems (e.g. MySQL/MariaDB SQL databases).

We first compare the memory demand for each of the 3 `Spectra` objects:

```{r}
#' Compare memory footprint.
print(object.size(s_mem), units = "MB")
print(object.size(s_mzr), units = "MB")
print(object.size(s_db), units = "MB")
```

As expected, the size of the `Spectra` object with the `MsBackendMemory` backend
is the largest, while those for the two other backends are much lower, with each
of them needing less than 1MB of total memory. Since the `MsBackendOfflineSql`
backend keeps only the primary keys of the spectra in memory it's memory
requirements are particularly small and is thus ideal to represent very large MS
experiments.

We next evaluate the performance to extract spectra variables from the 3
backends. We compare the time to extract the retention times from the 3
`Spectra` object using the `microbenchmark` function. In addition we disable any
parallel processing using `register(SerialParam())`.

```{r}
#' Compare performance to extract the retention times.
register(SerialParam())
library(microbenchmark)
microbenchmark(
    rtime(s_mem),
    rtime(s_mzr),
    rtime(s_db),
    times = 13
)
```

The `Spectra` object with the `MsBackendMemory` backend shows the highest
performance, followed by the `Spectra` object with the `MsBackendMzR`, while
extraction of spectra variables is slowest for the `Spectra` with the
`MsBackendOfflineSql`.

We next also evaluate the performance to extract peaks data, i.e. the individual
*m/z* and intensity values for all spectra in the two files.

```{r}
#' Compare performance to extract the peaks data.
microbenchmark(
    peaksData(s_mem),
    peaksData(s_mzr),
    peaksData(s_db),
    times = 7
)

```

The `MsBackendMemory` outperforms the two other backends also in this
comparison. Both the `MsBackendMzR` and `MsBackendOfflineSql` need to import the
data first into memory, the `MsBackendMzR` from the original mzML files and the
`MsBackendOfflineSql` from the SQL database.

At last we evaluate the performance to subset any of the 3 `Spectra` objects to
10 random spectra.

```{r}
#' Compare performance to subset the Spectra.
idx <- sample(seq_along(s_mem), 10)
microbenchmark(
    s_mem[idx],
    s_mzr[idx],
    s_db[idx]
)
```

Here, the `MsBackendOfflineSql` has a clear advantage over the two other
backends, because it needs to simply subset the integer vector of primary keys,
while for the `MsBackendMemory` we need to subset the whole data, and for the
`MsBackendMzR` the data frame with the spectra variables.


### Performance considerations and tweaks

Keeping all data in memory (e.g. by using a `MsBackendMemory`) has its obvious
performance advantages, but might not be possible for all MS
experiments. *On-disk* backends such as the `MsBackendMzR` or
`MsBackendOfflineSql` allow, due to their low memory footprint, to analyze also
very large data sets. This is further optimized by the build-in option for most
`Spectra` methods to load and process MS data in chunks in contrast to loading
and processing the full data as a whole. Performance of the `MsBackendMzR` and
`MsBackendSql`/`MsBackendOfflineSql` backends depends also on the I/O
capabilities of the hard disks containing the data files as well as the
configuration of the SQL database system used.


## Data manipulation and the lazy evaluation queue

Different backends can be helpful, but for some data representations changing or
adding values might simply not be possible. The `MsBackendMzR` backend for
example retrieves the peaks data directly from the raw data files and we don't
want any manipulation of *m/z* or intensity values to be directly written back
to these original data files. Also other backends, specifically those that
provide access to spectral reference databases (such as the
`MsBackendMassbankSql` from the `r Biocpkg("MsBackendMassbank")` package or the
`MsBackendCompDb` from the `r Biocpkg("CompoundDb")` package), are not supposed
to allow changes to be written back to these databases.

Any type of analysis requires however some sort of data manipulation, thus, we
need some sort of mechanism to *cache* changes to the data. Changes to spectra
variables are expected to be supported by the backend class itself.

To illustrate how this is implemented for our 3 backend classes we below add a
new spectra variable `"new_variable"` to each of our 3 `Spectra` objects.

```{r}
#' Assign new spectra variables.
s_mem$new_variable <- seq_along(s_mem)
s_mzr$new_variable <- seq_along(s_mem)
s_db$new_variable <- seq_along(s_mem)
```

A new spectra variable was now added to each of the 3 `Spectra` objects:

```{r}
#' Show spectra variables for one of the Spectra
spectraVariables(s_mzr)
```

The `MsBackendMemory` stores all spectra variables in a `data.frame` within the
object. The new spectra variable was thus simply added as a new column to this
`data.frame`.

**Warning**: **NEVER** access or manipulate any of the slots of a backend class
directly like below as this can easily result in data corruption.

```{r}
#' Direct access to the backend's data.
s_mem@backend@spectraData |> head()
```

Similarly, also the `MsBackendMzR` stores data for spectra variables (but not
peaks variables!) within a `DataFrame`. *Note*: the use of a `DataFrame` instead
of a `data.frame` in `MsBackendMzR` explains most of the performance differences
to subset or access spectra variables we've seen between this backend and the
`MsBackendMemory` above.

```{r}
s_mzr@backend@spectraData
```

Finally, also for the `MsBackendOfflineSql` backend, the new spectra variable is
cached within a `data.frame` inside the object.

```{r}
s_db@backend@localData |> head()
```

As a side effect, this can also be used for `MsBackendOfflineSql` backends to
cache spectra variables in memory for faster access.

```{r}
s_db2 <- s_db

#' Cache an existing spectra variable in memory
s_db2$rtime <- rtime(s_db2)

microbenchmark(
    rtime(s_db),
    rtime(s_db2),
    times = 17
)
```

*Note*: the base `MsBackendCached` class defined in the *Spectra* package
provides a caching mechanism for spectra variables. Thus, any `MsBackend`
implementation extending this base class automatically inherit this mechanism
from it and don't need to implement their own. Examples of backends extending
`MsBackendCached` are, among others, `MsBackendSql`, `MsBackendOfflineSql`,
and `MsBackendMassbankSql`.

Analysis of MS data requires, in addition to changing or adding spectra
variables as shown above also the possibility to manipulate peaks data (i.e. the
*m/z* and intensity values of the individual mass peaks of the spectra). As a
simple example we remove below all mass peaks with an intensity below 100 from
the data set and compare the number of peaks per spectra before and after that
operation.

```{r}
#' Remove all peaks with an intensity below 100
s_mem <- filterIntensity(s_mem, intensity = 100)

#' Compare the number of peaks before/after
boxplot(list(before = lengths(s_mzr), after = lengths(s_mem)),
        ylab = "number of peaks")
```

This operation did reduce the number of peaks considerably. We repeat this
operation now also for the two other backends.

```{r}
#' Repeast filtering for the two other Spectra objects.
s_mzr <- filterIntensity(s_mzr, intensity = 100)
s_db <- filterIntensity(s_db, intensity = 100)
```

And in fact the number of peaks were also reduced for these two backends even if
they do not keep the peak data in memory and, as discussed before, we do not
allow changes to the data file (or the database) containing the original data.

```{r}
#' Evaluate that subsetting worked for all.
median(lengths(s_mem))
median(lengths(s_mzr))
median(lengths(s_db))
```

Due to their (generally) limited and manageable size, the caching mechanisms
described above work well and efficiently for spectra variables. MS peaks data
(*m/z* and intensity values of the individual peaks of MS spectra) in contrast
represents however a much larger data volumes. Thus, in-memory caching of
changes to peaks data is not generally possible for all types of MS data.

For that reason we are not caching changes to the values within the objects, but
the actual data manipulation operation itself. This is done automatically by the
`Spectra` object (not the `MsBackend`!) for any of its (peaks) data manipulation
functions. Internally, they are stored in a *lazy processing queue*.

```{r}
#' Show the processing queue.
s_db@processingQueue
```

Each time peaks data is accessed (like with the `intensity` call in the example
below), the `Spectra` object will first get the *raw* peaks data from the
backend, check its processing queue and, if that is not empty, apply each of the
contained cached processing steps to the peaks data before returning it.

```{r}
#' Access intensity values and extract those of the 1st spectrum.
intensity(s_db)[[1L]]
```

As a nice side effect, if only a subset of the data is accessed, the data
manipulations need only to be applied to the requested data subset, and not the
full data, which can be beneficial for the performance.

```{r}
#' Compare extraction of intensity values.
microbenchmark(
    intensity(s_db[1:10]),
    intensity(s_db)[1:10],
    times = 7)
```

This mechanism is implemented in the `Spectra` object and is thus independent of
the backend. Ultimately, it allows to apply also data manipulations to MS data
from data resources that would be inherently *read-only* (such as e.g. reference
libraries of fragment spectra).

In addition, since we are not changing peaks data, we can also *undo* data
manipulations. A simple `reset` call on a `Spectra` object will restore the data
in the object to its initial state (in fact it simply clears the processing
queue).

```{r}
#' Restore the Spectra object to its initial state.
s_db <- reset(s_db)
median(lengths(s_db))
```


## Implementing your own `MsBackend`

- New backends for `Spectra` should extend the `MsBackend` class and implement
  the methods defined by its infrastructure.
- Extending other classes, such as the `MsBackendMemory` or `MsBackendCached`
  can help reducing the number of methods or concepts needed to implement.
- A detailed example and description is provided in the [*Creating new
  `MsBackend` classes*](https://rformassspectrometry.github.io/Spectra....)
  vignette - otherwise: open an issue on the [*Spectra* github
  repo](https://github.com/RforMassSpectrometry/Spectra).
- Add the following code to your *testthat.R* file to ensure compliance with the
  `MsBackend` definition:
  ```
  test_suite <- system.file("test_backends", "test_MsBackend",
                          package = "Spectra")

  #' Assign an instance of the developed `MsBackend` to variable `be`.
  be <- my_be
  test_dir(test_suite, stop_on_failure = TRUE)
  ```

## Final words

- The `r Biocpkg("Spectra")` package provides a powerful infrastructure to
  handle and process MS data in R:
  - designed to support very large data sets.
  - easily extendable.
  - support for a variety of MS data file formats and data representations.


# References
